\textbf{DiracDec C++/Coq Implementation\cite{ExampleCitation}}

### 2025-01-03 main.tex
update 2025/1/3

\section{Consideration}
\begin{itemize}
    \item AC equivalence and alpha equivalence.
    \item Efficient AC equivalence checking by sorting on syntax.
    \item Sorting requires that alpha equivalent terms have the same syntax.
    \item de Bruijn expression satisfies this requirement, but the typing and substitution becomes very complicated in the type index scenario.
    \item only compute de Bruijn form in the equivalence checking and sorting phase.
    \item pipeline: 
        \begin{enumerate}
            \item preprocessing
            \item 1st rewritings
            \item variable expansion
            \item 2nd rewritings 
            \item computing de Bruijn and sorting
        \end{enumerate}

@@

\section{TODO}
\begin{itemize}
    \item implemented trace output for the whole pipeline.
    \item better parser and output.
    \item add Mathematica support for scalars
    \item output trace to Coq
    \item consider labelled Dirac notations
    \item better trace output
    \item better error logic
    \item better format output
\end{itemize}

\section{Things to Note}
\begin{itemize}
    \item The context should also be maintained during rewriting matching.
    \item We don't allow eta reduction. It will intertwine with SUM-SWAP and break the confluence.
\end{itemize}

@@

\section{Language Syntax}
\begin{definition}[syntax]
    The syntax for type indices are defined as
    \begin{align*}
        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2\ |\ \{c_0, c_1, \cdots, c_n\}.
    \end{align*}
    The syntax for Dirac notation types is defined as 
    \begin{align*}
        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall x.T\ |\ \SET(\sigma). \\
    \end{align*}
    The syntax for Dirac notation terms is defined as
    \begin{align*}
        e ::=\ & x\ |\ (e_1, e_2)\ |\ \lambda x : T.e\ |\ \mu x.e\ |\ e_1\ e_2 \\
        & |\ 0\ |\ 1\ |\ \ADDS(e_1 \cdots e_n) \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ |\ \DOT(e_1\ e_2) \\
        & |\ \ZEROK(\sigma)\ |\ \ZEROB(\sigma)\ |\ \ZEROO(\sigma_1, \sigma_2)\ |\ \ONEO(\sigma) \\
        & |\ \ket{e}\ |\ \bra{t}\ |\ e^\dagger\ |\ e_1.e_2\ |\ \ADD(e_1 \cdots e_n)\ |\ e_1 \otimes e_2 \\
        & |\ \MULK(e_1\ e_2)\ |\ \MULB(e_1\ e_2)\ |\ \OUTER(e_1\ e_2)\ |\ \MULO(e_1\ e_2) \\
        & |\ \mathbf{U}(e)\ |\ e_1 \star e_2\ |\ \sum_{e_1} e_2.
    \end{align*}
\end{definition}
 % after consideration I still think keep DOT as binary is the best choice. It reduces the unstability of complicated matching in sequences.
Here $i$ is a natural number and $\$i$ represents the $i$-th bound variable in de Bruijn notation. 
Compared to [?], this syntax for Dirac notations merges the symbols with overlapped properties, such as the addition and scaling symbols for ket, bra and operator.
Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
There are five kinds of linear algebra multiplications among ket, bra and operator, whose properties are similar but still diverge to some extent. For example, the rules $(O_1 \cdot O_2) \cdot K\ \reduce\ O_1 \cdot (O_2 \cdot K)$ and $B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2$ indicate that the sorting of multiplication sequences depends on the subterm types. To avoid frequent but unnecessary type checkings, we encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$. They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.

@@

\begin{figure}
    \begin{gather*}
        \textbf{Index-Var} \qquad
        \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
        \qquad \qquad
        \textbf{Index-Prod} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index} \\
        \\
        \textbf{Index-Concrete} \qquad
        \frac{\WF(E)[\Gamma] \qquad c_i \notin E[\Gamma] \text{ for all } i \qquad c_i \text{ are unique}}{E[\Gamma] \vdash \{c_1, c_2, \cdots, c_n\} : \Index}
    \end{gather*}
    \caption{Rules for type index.}
\end{figure}

@@

\begin{ruletable}{Rules for addition and index in sum}
    %
    R-SUM-ADDS0
    & $\sum_{i \in M}(a_1 + \cdots + a_n) \ \reduce\ (\sum_{i \in M} a_1) + \cdots + (\sum_{i \in M} a_n) $ \\
    %
    R-SUM-ADD0
    & $\sum_{i \in M}(X_1 + \cdots + X_n) \ \reduce\ (\sum_{i \in M} X_1) + \cdots + (\sum_{i \in M} X_n) $ \\
    %
    R-SUM-ADD1
    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}X + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+1).X $ \\
    %
    R-SUM-ADD2
    & $Y_1 + \cdots + \sum_{i \in M}X + \cdots + \sum_{i \in M}(a.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(1+a).X $ \\
    %
    R-SUM-ADD3
    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}(b.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+b).X $ \\
    %
    R-SUM-INDEX0
    & $ \sum_{i \in \mathbf{U}(\sigma \times \tau)} A \ \reduce\ \sum_{j \in \mathbf{U}(\sigma)} \sum_{k \in \mathbf{U}(\tau)} A\{i/(j, k)\} $ \\
    %
    R-SUM-INDEX1
    & $ \sum_{i \in M_1 \star M_2} A \ \reduce\ \sum_{j \in M_1} \sum_{k \in M_2} A\{i/(j, k)\} $ \\
    %
    R-SUM-SWAP
    & $ M_1 < M_2 \vdash \sum_{i \in M_2} \sum_{j \in M_1} X \ \reduce\ \sum_{j \in M_1} \sum_{i \in M_2} X $
\end{ruletable}

>>>>

\section{Consideration}
\begin{itemize}
    \item We can emphasize that we are moving from pure term rewriting to a hybrid algorithm.
    \item AC equivalence and alpha equivalence.
    \item Efficient AC equivalence checking by sorting on syntax.
    \item Sorting requires that alpha equivalent terms have the same syntax.
    \item de Bruijn expression satisfies this requirement, but the typing and substitution becomes very complicated in the type index scenario.
    \item only compute de Bruijn form in the equivalence checking and sorting phase.
    \item pipeline: 
        \begin{enumerate}
            \item preprocessing
            \item rename unique bound variable names
            \item 1st rewritings
            \item variable expansion
            \item 2nd rewritings 
            \item sorting modulo bound variables
            \item sorting successive sum. The order depends on there occurances in the last sorting result. If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
            \item computing de Bruijn
        \end{enumerate}

@@

\section{TODO}
\begin{itemize}
    \item implemented trace output for the whole pipeline.
    \item better output.
    \item add Mathematica support for scalars
    \item output trace to Coq
    \item consider labelled Dirac notations
    \item better trace output
    \item better error logic
    \item better format output
    \item remove term bank and hash consing
\end{itemize}

\section{Things to Note}
\begin{itemize}
    \item The context should also be maintained during rewriting matching.
    \item We don't allow eta reduction. It will intertwine with SUM-SWAP and break the confluence.
    \item In each Delta reduction, the bound variables are replaced with unique variables. This should help solve the problem of conflicting variable names during substitution.
    \item I guess it may still be necessary to try all different bound variable assignments.
\end{itemize}

@@

\section{Language Syntax}
\begin{definition}[syntax]
    The syntax for type indices are defined as
    \begin{align*}
        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2\ |\ \mathsf{Qdit}_n.
    \end{align*}
    The syntax for Dirac notation types is defined as 
    \begin{align*}
        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall x.T\ |\ \SET(\sigma). \\
    \end{align*}
    The syntax for Dirac notation terms is defined as
    \begin{align*}
        e ::=\ & x\ |\ (e_1, e_2)\ |\ \lambda x : T.e\ |\ \mu x.e\ |\ e_1\ e_2 \\
        & |\ 0\ |\ 1\ |\ \ADDS(e_1 \cdots e_n) \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ |\ \DOT(e_1\ e_2) \\
        & |\ \ZEROK(\sigma)\ |\ \ZEROB(\sigma)\ |\ \ZEROO(\sigma_1, \sigma_2)\ |\ \ONEO(\sigma) \\
        & |\ \ket{e}\ |\ \bra{t}\ |\ e^\dagger\ |\ e_1.e_2\ |\ \ADD(e_1 \cdots e_n)\ |\ e_1 \otimes e_2 \\
        & |\ \MULK(e_1\ e_2)\ |\ \MULB(e_1\ e_2)\ |\ \OUTER(e_1\ e_2)\ |\ \MULO(e_1\ e_2) \\
        & |\ \mathbf{U}(e)\ |\ e_1 \star e_2\ |\ \sum_{e_1} e_2.
    \end{align*}
\end{definition}
 % after consideration I still think keep DOT as binary is the best choice. It reduces the unstability of complicated matching in sequences.
Here $i$ is a natural number and $\$i$ represents the $i$-th bound variable in de Bruijn notation. 
Compared to [?], this syntax for Dirac notations merges the symbols with overlapped properties, such as the addition and scaling symbols for ket, bra and operator.
Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
There are five kinds of linear algebra multiplications among ket, bra and operator, whose properties are similar but still diverge to some extent. For example, the rules $(O_1 \cdot O_2) \cdot K\ \reduce\ O_1 \cdot (O_2 \cdot K)$ and $B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2$ indicate that the sorting of multiplication sequences depends on the subterm types. To avoid frequent but unnecessary type checkings, we encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$. They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.

@@

\begin{figure}
    \begin{gather*}
        \textbf{Index-Var} \qquad
        \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
        \qquad \qquad
        \textbf{Index-Prod} \qquad
        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index} \\
        \\
        \textbf{Index-Qudit} \qquad
        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash \mathsf{Qudit}_n : \Index}
    \end{gather*}
    \caption{Rules for type index.}
\end{figure}

@@

\begin{ruletable}{Rules for addition and index in sum}
    %
    R-SUM-ADDS0
    & $\sum_{i \in M}(a_1 + \cdots + a_n) \ \reduce\ (\sum_{i \in M} a_1) + \cdots + (\sum_{i \in M} a_n) $ \\
    %
    R-SUM-ADD0
    & $\sum_{i \in M}(X_1 + \cdots + X_n) \ \reduce\ (\sum_{i \in M} X_1) + \cdots + (\sum_{i \in M} X_n) $ \\
    %
    R-SUM-ADD1
    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}X + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+1).X $ \\
    %
    R-SUM-ADD2
    & $Y_1 + \cdots + \sum_{i \in M}X + \cdots + \sum_{i \in M}(a.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(1+a).X $ \\
    %
    R-SUM-ADD3
    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}(b.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+b).X $ \\
    %
    R-SUM-INDEX0
    & $ \sum_{i \in \mathbf{U}(\sigma \times \tau)} A \ \reduce\ \sum_{j \in \mathbf{U}(\sigma)} \sum_{k \in \mathbf{U}(\tau)} A\{i/(j, k)\} $ \\
    %
    R-SUM-INDEX1
    & $ \sum_{i \in M_1 \star M_2} A \ \reduce\ \sum_{j \in M_1} \sum_{k \in M_2} A\{i/(j, k)\} $
\end{ruletable}
